---
title: "Joins"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Joins are essential to queries involving data that spans multiple tables. Note that the **cross join** is not covered. 

## Sample Database

```{r}
library(RSQLite)
db <- dbConnect(SQLite(), ":memory:")
```
```{sql connection = db}
/* foreign key is off to allow dangling references */
PRAGMA foreign_keys = OFF
```

```{sql connection = db}
create table office (
 oid integer primary key,
 num text
);
```
```{sql connection = db}
create table empl (
 eid integer primary key,
 name text,
 oid integer,
 foreign key (oid) references office(oid)
);
```

Note that some employees do not have an office, and vice versa. 

```{sql connection = db}
insert into office values 
 (10,"NI 132F"),
 (20,"WVH 310A"),
 (30,"RY 611"),
 (40,"CH 103"),
 (50,"106A");
```
```{sql connection = db}
insert into empl values 
 (601,"Jeff Goldblum",10),
 (602,"Ann Hathaway",20),
 (603,"Michael Keaton",30),
 (604,"Jennifer Hudson",NULL),
 (605,"Mark Wahlberg",44),
 (609,"Helen Miren", 50);
```

## Cartesian Product

When query contains 2+ tables in *FROM*, SQL produces Cartesian Product of the tables, i.e. all combinations of rows 

* note that foreign key and primary key only match on some rows 

```{sql connection = db}
select * from empl, office;
```

<br>

## Inner Join

Inner join is the most common form of join - joins rows of two tables such that PK in one table matches FK in other.

* **Standard WHERE Clause with PK/FK Matching**

  Simple and common, but not as obvious as other methods.
  
  ```{sql connection = db}
  select e.name as EmployeeName, o.num as OfficeNum
    from empl as e, office as o
   where e.oid = o.oid;
  ```
  
* **Inner Join Syntax**

  Preferred method, clearly expresses join and separates join clause from other selection criteria. Note that `inner` keyword is optional. 
  
  ```{sql connection = db}
  select *
    from empl e inner join office o on (e.oid = o.oid);
  ```
  
* **Natural Join**
  
  A natural join is an inner join where the database matches tables based on common column names. 
  
  * This won't work as expected if any columns other than the PK/FK columns have matching names. 
  
  ```{sql connection = db}
  select * from empl natural join office;
  ```
  
## Joining Multiple tables 

Joining multiple tables is achieved by sequentially joining pairs of tables. All tables must be connected by some path to any other table. 

```{sql connection=db}
create table campus (
 cid text primary key,
 name text not null,
 city text not null,
 state text not null,
 country text not null
);
```
```{sql connection=db}
alter table office
  add column cid text;
```
```{sql connection=db}
insert into campus values 
 ('BOS', 'Main Campus Boston', 'Boston', 'MA', 'USA'),
 ('SV', 'Silicon Valley', 'San Jose', 'CA', 'USA'),
 ('TOR', 'Toronto', 'Toronto', 'ON', 'Canada'),
 ('SEA', 'Seattle', 'Seattle', 'WA', 'USA');
```
```{sql connection=db}
update office
   set cid = 'BOS'
 where oid IN (10, 20, 30, 40);
```
```{sql connection=db}
update office
   set cid = 'SEA'
 where oid IN (50);
```

Now, to join, we can use WHERE clause or a join statement like above. 

```{sql eval = FALSE}
/* not evaluated */
select e.name, o.num, c.name
  from empl e, office o, campus c
 where e.oid = o.oid and o.cid = c.cid;
```
```{sql connection = db}
select e.name, o.num, c.name
  from empl e join office o on (e.oid = o.oid)
              join campus c on (o.cid = c.cid);
```

## Outer Joins 

Outer join has three different types, each of which add a specified set of unmatched rows. 

* **Left Outer Join**
  
  Left outer joins select all common rows plus all unmatched rows from the left table in join specification. Unmatched rows have missing information populated with NA. 
  
  ```{sql connection = db}
  select *
    from empl e left join office o on (e.oid = o.oid)
  ```
  
  Note that this can be used to find unmatched rows and missing fks, which usually can't happen but might with low-quality imported data. 
  
  ```{sql connection = db}
  /* find missing FK's and rows with missing offices */
  select *
    from empl e left join office o on (e.oid = o.oid)
  except
  select *
    from empl e inner join office o on (e.oid = o.oid);
  ```
  ```{sql connection = db}
  /* find dangling references */
  select e.name, e.oid
    from empl e left join office o on (e.oid = o.oid)
  except
  select e.name, e.oid
    from empl e inner join office o on (e.oid = o.oid)
  except
  select e.name, e.oid
    from empl e
   where e.oid is null;
  ```

* **Right Outer Join**

  This is the same as a left outer join except that unmatched rows come from table on the right. This isn't explicitly supported, so just reverse a left join. 
  
  ```{sql connection = db}
  select * from office o left join empl e on (e.oid = o.oid);
  ```

* **Right Outer Join**

  Full outer join is the union of a left and right outer join, showing all matching rows as well as unmatched rows from the right and left table. 
  
  * For this to work, column order must be explicitly specified: `*` doesn't work as queries return columns in different orders. 

  ```{sql connection = db}
  select e.name, o.num
    from office o left join empl e on (e.oid = o.oid)
  union
  select e.name, o.num
    from empl e left join office o on (e.oid = o.oid)
  ```

## Anti-Join

In an anti-join we're looking to find all rows that are in one table, but not in the other. To do this, do a left join and then filter out anything where the FK is NULL. 

```{sql connection = db}
select o.oid, o.cid
  from office as o
  left join empl as e on (o.oid = e.oid)
 where e.oid = null;
```

## Self Joins

self joins can be useful on associative join tables to determine membership to multiple groups. 

```{sql eval = FALSE}
/* determine which students were assigned to both Teams A and B */
select * 
  from StudentTeamAssignment a1
  join StudentTeamAssignment a2 on a1.StudentFK = a2.StudentFK
 where a1.TeamFK = 'A' AND a2.TeamFK = 'B';
```

```{r include = FALSE}
dbDisconnect(db)
```










