---
title: "SELECT"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(RSQLite)

db <- dbConnect(SQLite(), "AssetDB.db")
```

<br>

## Query Basic Commands

Queries are issued through select statements. 

* only `SELECT` and `FROM` are required clauses.

```{sql eval = FALSE}
SELECT [columns]
  FROM [tables]
 WHERE [column conditions]
 GROUP BY [column]
   HAVING [group criteria]
 ORDER BY [columns];
```

Retrieve columns by listing their names, or use * to retrieve them all. 

```{sql connection = db}
SELECT name, email FROM Author;
```

### `AS`

Use `AS` to create **column aliases** and rename retrieved variables. 

* most useful for reporting or simplying conditions for `WHERE`. 

```{sql connection = db}
select name as FullName, email as EMail from Author;
```

### `LIMIT`

equivalent to `head()` - limits result to first *n* rows. 
```{sql connection = db}
select name, email from author limit 2;
```

### `ORDER BY`

Results are returned in a table known as **result set**. These results do **not** have a specific order. 

Specify order of rows with `ORDER BY`

* Supply additional column rows to break ties
* Specify sorting direction as ASC or DESC
* Text is sorted alphabetically

```{sql connection = db}
  select name, email 
    from Author
order by name, email desc;
```

### `DISTINCT`

Use `DISTINCT` to eliminate duplicate rows. It applies to all selected columns. 

```{sql eval = FALSE}
select distinct district from Address;
```

<br>

## Query Arguments

Queries can contain literals, expressions, and various methods. 

**Literals** can select only values matching that literal. 

```{sql connection = db}
select 'Joe Chang' name, email from Author;
```

**Expressions** can manipulate returned values

```{sql eval = FALSE}
/* example: not evaluated */
select title, length/60.0 AS hrs from Film;
```

**Functions** can perform operations on returned values

* inbuilt methods include `ABS`, `POWER`, `FLOOR`, `LENGTH(string)`, `SUBSTR(s, m [,n])`, `ADD_MONTHS(date, n)`, etc. 

```{sql connection = db}
select upper(name), email from Author;
```

```{r}
dbDisconnect(db)
```