---
title: "Triggers assignment - Omi Johnson"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

## Table Setup (prior assignment) 

```{r initialization}
# Initialization
library(RSQLite)
dbcon <- dbConnect(RSQLite::SQLite(), "OJohnson_db.db")
```
```{sql connection = dbcon}
/* Enabling foreign key checking */
PRAGMA foreign_keys = ON
```
```{sql connection = dbcon}
/* Creating Module Table */
/* number attribute renamed to mnumber to avoid overlap with reserved keyword. */
CREATE TABLE Module (
  mnumber TEXT NOT NULL, 
  title TEXT NOT NULL, 
  lengthInMin INTEGER NOT NULL,
  CONSTRAINT num_is_pk PRIMARY KEY (mnumber)
)
```
```{sql connection = dbcon}
/* Creating lookup table for author affiliation -
  choice to use name itself as primary key was deliberate choice to 
  improve clarity and ease of information access.
*/
CREATE TABLE Affiliation (
  atype TEXT NOT NULL, 
  CONSTRAINT at_is_pk PRIMARY KEY (atype)
);
```
```{sql connection = dbcon}
/* Creating Author Table */
CREATE TABLE Author (
  aid INTEGER NOT NULL,
  name TEXT NOT NULL,
  email TEXT NOT NULL, 
  affiliation TEXT NOT NULL,
  CONSTRAINT aid_is_pk PRIMARY KEY (aid),
  CONSTRAINT aff_fk_aff FOREIGN KEY (affiliation) REFERENCES Affiliation(atype)
);
```
```{sql connection = dbcon}
/* Creating Authorship junction Table */
CREATE TABLE Authorship (
  aid INTEGER NOT NULL,
  mnumber TEXT NOT NULL, 
  CONSTRAINT all_is_pk PRIMARY KEY (aid, mnumber),
  CONSTRAINT aid_fk_au FOREIGN KEY (aid) REFERENCES Author(aid),
  CONSTRAINT mnum_fk_af FOREIGN KEY (mnumber) REFERENCES Module(mnumber)
)
```
```{sql connection=dbcon}
INSERT INTO Affiliation (atype) VALUES
  ("PT"), ("FT"), ("Contract");
```
```{sql connection=dbcon}
INSERT INTO Author (aid, name, email, affiliation) VALUES
  (14, "John", "john@email.com", "FT"),
  (15, "Ash", "ash@email.com", "PT"),
  (16, "Doe", "doe@email.com", "PT"),
  (17, "Diane", "diane@email.com", "Contract");
```
```{sql connection=dbcon}
INSERT INTO Module (mnumber, title, lengthInMin) VALUES
  ("01", "Dogs in Nature", 12),
  ("02", "Fall leaves", 7),
  ("03", "Party planning", 90);
```
```{sql connection=dbcon}
INSERT INTO Authorship (aid, mnumber) VALUES
  (14, "01"),
  (14, "02"),
  (17, "03"),
  (15, "02");
```

<br>

## Additional Table Insertion

The `employmentChange` table logs changes in an employment for an author. 

```{sql connection = dbcon}
/* assuming that employment change cannot happen twice within a day */
CREATE TABLE IF NOT EXISTS employmentChange (
  aid INTEGER NOT NULL,
  prior_aff TEXT,
  new_aff TEXT,
  changeDate DATE NOT NULL,
  CONSTRAINT aid_date_is_pk PRIMARY KEY (aid, changeDate),
  CONSTRAINT aid_ref_auth FOREIGN KEY (aid) REFERENCES Author(aid),
  CONSTRAINT p_aff_ref_aff FOREIGN KEY (prior_aff) REFERENCES Affiliation(atype),
  CONSTRAINT n_aff_ref_aff FOREIGN KEY (new_aff) REFERENCES Affiliation(atype)
)
```

<br>

## Trigger Creation and Explanation

The `logEmploymentChange` trigger logs changes in employment status to the `employmentChange` table in the case an author's employment type changed. It records the current date, the prior and new employment type, and the author id to the `employmentChange` table. This trigger is important because keeping track of employment type history for authors, and the dates for each switch, is critical information for a company to keep an accurate log of individual author's work history and to derive information on the number of employees within a specific type at a given point in time.

```{sql connection = dbcon}
DROP TRIGGER IF EXISTS logEmploymentChange
```

```{sql connection = dbcon}
CREATE TRIGGER logEmploymentChange 
  AFTER UPDATE ON Author
  FOR EACH ROW 
  WHEN (NEW.Affiliation <> OLD.Affiliation)
    BEGIN
      INSERT INTO employmentChange VALUES
      (NEW.aid, OLD.Affiliation, NEW.Affiliation, DATE('now'));
END;
```

<br>

## Trigger testing 

```{sql connection = dbcon}
/* does logEmploymentChange work correctly moving from one role to another? Does it ignore 
cases where the prior and updated employment type are the same? */
UPDATE Author 
   SET Affiliation = "FT"
 WHERE aid IN (14, 15, 17)
```
```{sql connection = dbcon}
/* Does logEmploymentChange ignore cases where a different attribute is updated? */
UPDATE Author 
   SET email = "doe12@email.com"
 WHERE aid IN (16)
```
```{sql connection = dbcon}
/* Does logEmploymentChange log employment changes when multiple attributes update? */
UPDATE Author 
   SET email = "doe21@email.com", Affiliation = "Contract"
 WHERE aid IN (16)
```

```{sql connection = dbcon}
/* expect three rows */
SELECT * FROM employmentChange
```

```{r}
dbDisconnect(dbcon)
```

```{r eval = FALSE}
file.remove("OJohnson_db.db") #run manually to allow re-execution
```

<br>