---
title: "Parsing XML with R"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Parse XML with R lesson ](http://artificium.us/lessons/06.r/l-6-114-parse-xml-r-primer/l-6-114.html)
[Reading XML into SQLite](http://artificium.us/lessons/06.r/l-6-328-xml-to-reldb-sqlite/l-6-328.html)

Many of the concepts here extend to other programming languages. 

```{r}
library(XML) # there are other options!
library(stringr)
```

# Loading an XML File

XML files can be loaded and parsed as both a local file or from a URL. 

```{r eval = FALSE}
xmlFile <- "BookCatalog.xml"
xmlURL <- "http://artificium.us/lessons/06.r/l-6-114-parse-xml-r-primer/BookCatalog.xml"

domB <- xmlParse(xmlFile)
```

`xmlParse()` returns a reference to an internal tree of nodes representing the DOM for the XML.

  * `xmlTreeParse()` also returns a DOM, but represented as an R data structure rather than C data structure. It allows access using the $ operator, but in practice `xmlParse()` is used due to its better efficiency. 

## Validation

`xmlParse()` doesn't validate the file against a DTD or XSD - it only checks if the file is well-formed. To ensure that te SML conforms to a DTD, use `validate = T`. 

Other parameters include `trim` (strip whitespace), `getDTD` (should DTD be returned?), and `isURL`. 

```{r eval = FALSE}
domP <- xmlParse("pagevisits.xml", validate = T)
```

<br>

# Extracting Elements

There are two common ways to extract element values -- through indexing the DOM and through XPath. 

We use an example XML: 

```xml
<catalog>
   <book id="bk101" edition="3">
      <author>
        <surname>Gambardella</surname>
        <given>Matthew</given>
      </author>
      <title>XML Developer's Guide</title>
      <outofprint />
      <price currency="R$">349</price>
   </book>
   <book id="bk102" edition="1">
      <author>
        <surname>Ralls</surname>
        <given>Kim</given></author>
      <title>Midnight Rain</title>
      <price currency="US$">5.95</price>
   </book>
</catalog>
```

We begin by loading the XML. 

```{r}
xmlDoc <- xmlParse("files/SimpleXML.xml", validate = F)
```

## Indexed Access

To access elements, we first acquire the root element of the DOM tree. 

```{r}
root <- xmlRoot(xmlDoc)
```

Accessing the *ith* child node underneath the root can be done using the list access operator `[[i]]`. This can be continued all the way down a tree. 

```{r}
print(root[[1]])

aNode <- root[[2]][[1]][[2]]

print(aNode)
```

To retrieve the value of a node, use `xmlValue()`

```{r}
xmlValue(aNode) |> print()
```

One advantage of this model is that the tree can be processed in a loop.

```{r}
# extracting all author surnames
n <- xmlSize(root)

names <- c(length = n)

for (i in 1:n) {
  names[i] <- xmlValue(root[[i]][[1]][[1]])
}

cat(names)
```

### Optional Elements

Optional elements can be tricky when using indexed access. One technique to handle this is by checking the name of a node using `xmlName()`. 

```{r}
node <- xmlName(root[[1]][[3]]) 
ifelse (node == "outofprint", "Out of Print", "Printing")
```

## Storing XML in Data Frame

Storing XML into a data frame is a common strategy when converting data from XML to a tabular format or saving to relational database

```{r}
# number of <book> nodes
n <- xmlSize(root)

# empty data frame
df <- data.frame(title = as.character(n),
                 price = as.numeric(n))

for (i in 1:n)
{
  title <- xmlValue(root[[i]][[2]])
  node <- xmlName(root[[i]][[3]])
  price.node <- ifelse(node == "outofprint", 
                       4, 3)
  
  price <- xmlValue(root[[i]][[price.node]])
  price.value <- as.numeric(price)
  
  # add to data frame
  df[i,"title"] <- title
  df[i,"price"] <- price.value
}

head(df,3)
```

### Extracting Attributes

Attributes are extracted using `xmlAttrs()`, which accepts a method and returns a list of all attributes. 

```{r}
xmlAttrs(root[[1]])
```

To extract a specific attribute, use `xmlGetAttr()`. 

```{r}
xmlGetAttr(root[[1]], "edition")
```


## XPath

[XPath in R](http://artificium.us/lessons/06.r/l-6-324-parse-xml-dom/l-6-324.html)

[Data retrieval with XPath Examples](http://artificium.us/lessons/06.r/l-6-303-xpath-in-r/l-6-303.html)

[XPath functions](https://northeastern.instructure.com/courses/123281/pages/30-min-read-a-primer-on-xpath-functions?module_item_id=7536375)

XPath expressions can access elements and attributes in a more elegant, simple way. 

`xpathSApply()` applies an XPath expression and returns a list of all nodes matching that expression.

* there is also `xpathApply()`, but this is not preferred as it returns a list rather than a simplified vector

```{r}
xpathSApply(xmlDoc, 
            "//book/title")
```

To retrieve values of elements, add `xmlValue()` as a parameter. We don't need to add xmlValue when retrieving attribute values. 

```{r}
xpathSApply(xmlDoc, 
            "//book/title", 
            xmlValue)
```

### Missing Elements

Check for optional elements using XPath by finding the length of the returned list.

```{r}
xpathSApply(xmlDoc, "//book/genre", xmlValue) |>
  length()
```

## XPath with Indexed Access

If there are multiple of an element within a parent, or none at all, combinations of XPath and indexed access may be necessary. 

Consider a more complex XML with multiple prices, where we only what to extract US prices. Here's an example of a node: 

```xml
<catalog>
   <book id="bk101" edition="3">
      <author>
        <surname>Gambardella</surname>
        <given>Matthew</given>
      </author>
      <title>XML Developer's Guide</title>
      <outofprint />
      <price currency="R$">349</price>
      <price currency="US$">29.95</price>
      <price currency="€">34.00</price>
   </book>
   ...
 </catalog>
```

There are multiple approaches to solve this, but one is below that affords more flexibility than pure XPath. 

```{r}
xmlDoc <- xmlParse("files/SimpleXML-2.xml")

n <- xmlSize(root)

# get all <book> nodes from the XML
books <- xpathSApply(xmlDoc, "//book")

# iterate over the <book> nodes
for (i in 1:n) {
  # get the i-th book node
  aBook <- books[[i]]
  
  # use XPath to extract the <price> child elements
  xpathSApply(aBook,
              "price[./@currency='US$']",
              xmlValue)
}
```

# Loading simple XML into a dataframe

[lesson link](http://artificium.us/lessons/06.r/l-6-323-load-xml-xmlToDataFrame/l-6-323.html)

Simple row/column XML files can be loaded directly into a dataframe using `xmlToDataFrame()`. It's convenient to load simple XML files that are often the result of data exports. 

```{r eval = FALSE}
df <- xmlToDataFrame(filename, 
                     stringsAsFactors = F)
```

Note that all values will be loaded as text and may require explicit numeric conversion. 

For more complex XML documents, a combination of XPath and DOM must be used. 


