---
title: "SQLite Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Client/Server databases**, aka server based: 

* applications first connect to TCP/IP port, use port to send SQL database requests and receive table results
* all database processing occurs in separate process, most likely on shared server computer 

<br>

## SQLite ##

SQLite is a relational database that is purposefully lightweight and ideal for learning and for small applications not requiring concurrent access. 

* self contained
* serverless 
* zero-configuration
* transactional 

**Serverless**: 

* database contained within single file on client application storage system
* database is single user, not shared
* SQL processing occur within compiled SQLite library 

## Connecting to SQLite ##

We can access the SQLite database through the RSQLite package.

```{r}
library(RSQLite)

# pass in name for database -- if doesn't already exist, new database is created 
dbcon <- dbConnect(RSQLite::SQLite(), "myDB.db")

dbcon
```

After establishing connection, can access database using `RSQLite` methods, eg. `dbExecute()`, `dbGetQuery()`, `dbSendStatement()`. See other pages for more information.


### Working with SQL in RMarkdown ###

SQL operations are done within a `{sql connection=dbcon}` chunk. Note the `connection` argument. 

* R functions also exist for these statements, but `{sql}` is simpler. 
* Semicolons are optional in an {sql} chunk, but good coding practice to include
* commands are case insensitive; variables are not
  * prefer uppercase
* **Only include one SQL command per chunk.**

```{sql connection=dbcon}
create table lessons (
  lname text not null,
  llength integer not null
)
```
```{sql connection=dbcon}
insert into lessons (lname, llength) values
  ('Intro C++', 90), 
  ('Intro Java', 90),
  ('SQL Joins', 60);
```
```{sql connection=dbcon} 
select * from lessons;
```

```{r, include = FALSE}
dbDisconnect(dbcon)

file.remove("myDB.db")
```





























