---
title: "Lookup Tables"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RSQLite)
dbcon <- dbConnect(RSQLite::SQLite(), "myDB.db")
```

**Lookup/Association tables** associate a value with other values, eg clothes sizing with measurements 

* each row in lookup table is unique, making entire row a primary key
  * generally no surrogate/synthetic key unless lookup values referenced from another table
* used to implement enumerated values/factors
* fixed - no manipulation

## Creating and Using Lookup Tables

First, create and populate the table. 

```{sql connection = dbcon}
create table if not exists sizeLookup (
  germanSize integer not null, 
  waist_cm integer not null,
  length_cm integer not null,
  us_waist_in integer not null,
  us_inseam_in integer not null
)
```
```{sql connection = dbcon}
insert into sizeLookup values 
  (46,81,82,31,32),
  (52,93,86,36,34),
  (25,93,80,36,30),
  (110,101,93,40,36)
```

Then, to search table we can use queries.

```{sql connection = dbcon}
select us_waist_in, us_inseam_in
  from sizeLookup
 where germanSize = 25;
```

## Enumerated Values 

Use for enumerated values is useful, especially when database does not naively support enumerations 

```{sql connection = dbcon}
create table if not exists roles (
  rid integer not null primary key,
  rclass text
)
```
```{sql connection = dbcon}
insert into roles values
  (1, "E"), (2, "D"), (3, "C"), (4, "B"), (5, "A"), (6, "S");
```
```{sql heroes, connection = dbcon}
create table if not exists heroes (
  hid integer not null primary key,
  hname text not null,
  hrole integer not null, 
  foreign key (hrole) references roles(rid)
);
```
```{sql connection = dbcon}
insert into heroes values 
  (13, "Mirko", 6), 
  (12, "Adam", 1), 
  (11, "Jimbei", 5)
```

Query to list members and their role:

```{sql connection = dbcon}
select h.hname as 'Name', r.rclass as 'Role'
  from heroes h inner join roles r on (h.hrole = r.rid);
```

Alternatively, can make the name of the enumeration itself if the primary key: 

* eliminates need for join in queries

```{sql connection = dbcon}
create table if not exists roles (
  rname text not null primary key
)
```

Enumerations are also supported with constraints. For SQLite, use the `check` constraint ~ other databases may have an `enum` type. 

* This prevents levels from being queried directly, therefore cannot be used to populate drop-down menus, etc.

```{sql connection = dbcon}
create table if not exists heroes (
  hid integer not null primary key,
  hname text not null,
  hrole integer not null,
  check (hrole in ('E', 'D', 'C', 'B', 'A', 'S'))
)
```

## Benefits ##

* **Storage Management**: prevent repitition of enumeration, saving storage
* **Improved performance**: lookup tables up performance because 'join' is done with key values
* **simple data modification**: changes to an enumerated value or conversion details only need to be made within the lookup table - no reference changes
* **faster search**: lookup tables are generally small, making it faster to search data 

```{r include = FALSE}
file.remove("myDB.db")
```