---
title: "Database Meta Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RSQLite)
dbcon <- dbConnect(RSQLite::SQLite(), "myDB.db")
```

Database meto info is stored in tables and can be queried -- application code therefore independent from hard coded schema definitions. 

All commands to query database meta data will work for both SQLite database and MySQL/etc. 

```{sql connection = dbcon}
create table plants (
  pname text not null,
  pleaves integer not null
);
```
```{sql connection = dbcon}
insert into plants values
  ("pilea", 30), 
  ("pothos", 30), 
  ("ming aralia", 100);
```

## Query Table Structure ##

To find the structure of a table:

* execute `.schema` or `.schema table_name` for specific table
* SQLite command from SQLite console
* `pragma` command from program
  ```{sql connection = dbcon}
  pragma table_info(plants);
  ```
  **Return values:** column names, attributes, null status, default value, whether pk 
* query tables containing table information, *ie* **metadata**. 
  ```{sql connection = dbcon}
  /* in sqlite, sqlite_schema contains table info */
  select * from sqlite_schema where name = 'plants';
  ```
* retrieve SQL command used to create table
  * use R block to capture SQL used 
  ```{r}
  rs <- dbGetQuery(dbcon, paste0(
    "select sql", 
    "  from sqlite_schema", 
    "  where name = 'plants'"))
  
  cat(rs[[1]])
  ```

## List Schema ##

To list names of all tables in database (omitting tables holding schema), use

```{sql connection = dbcon}
select name
  from sqlite_schema
 where `type` = 'table'
   and name not like 'sqlite_%';
```

## R DBI convenience methods ##

R has a collection of methods for querying db metainfo, such as `dbListTables(connection)` and `dbListFields(connection, "table_name")`. More on this [here](http://artificium.us/lessons/06.r/l-6-301-sqlite-from-r/l-6-301.html#Querying_Data_Frames_with_sqldf). 

```{r include = FALSE}
file.remove("myDB.db")
```