---
title: "Primary keys with auto_increment and rowid"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RSQLite)
dbcon <- dbConnect(RSQLite::SQLite(), "myDB.db")
```

```{sql connection = dbcon}
create table lessons (
  lname text not null,
  llength integer not null default 0);
```
```{sql connection=dbcon}
insert into lessons values
  ('Intro C++', 90), 
  ('Intro Java', 90),
  ('SQL Joins', 60);
```

## Row ID's as implicit primary keys ##

SQLite adds an automatic, hidden `rowid` column to all created tables. 

* eliminates requirement for explicit primary key column
* also accessible as `_rowid` and `oid`
* generated as integers starting at 1. will reuse deleted values. 

```{sql connection=dbcon}
select rowid, lname, llength from lessons;
```

use `without rowid` to create a table without a row id. In this case a primary key column must be specified explicitly

```{sql eval = FALSE}
create table lessons (
  lid integer primary key,
  lname text not null, 
  llength integer not null default 0,
) without rowid;
```

## Automatic PK's with `autoincrement`##

Use the `autoincrement` attribute to make SQLite generate a unique, increasing value for an element.

* SQLite only guarantees increasing value, not sequential 

```{sql colors, connection = dbcon}
create table colors (
  cid integer primary key autoincrement,
  cname text not null
);
```

Auto incremented variables do not require explicit declaration in `insert` commands. 

* Note that if omitting columns, must specify columns within declaration
* No values reused -- if rows are deleted, more likely to exceed max value

```{sql insert1, connection = dbcon}
insert into colors (cname) values
  ('Cyan'), ('Magenta');
```
 
Variables can still be explicitly declared, but affects future increments -- will increment from largest value

```{sql insert2, connection = dbcon}
insert into colors values 
  (433, 'Purple');
```
```{sql insert3, connection = dbcon}
insert into colors (cname) values
  ('Blue')
```
```{sql selectc, connection = dbcon}
select * from colors;
```

```{r include = FALSE}
file.remove("myDB.db")
```

## Takeaways ##

* application generated PK's are preferable
* rowid is potentially slow in generating keys -- tables should generally be created without
* PK's from autoincrement, rowid is convenient, but can mean slower performance and table relationship issues


