---
title: "DO: Query XML"
author: "Omi Johnson"
date: "2022-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(XML)

# reading in the XML
xmlDOM <- xmlParse("Books-v4.xml")

# helper function that runs an XPath expression
# ... allows for a large xpath expression to be broken across multiple lines
xpath_help <- function(...) {
  xpathSApply(xmlDOM, paste0(...))
}
```

## Part 1

What are the titles of all books selling for more than $50? List the titles.

```{r}
xpath_help("//book[price > 50]/title/text()")
```
Who is the author of the book with the title "Oberon's Revenge"?

```{r}
xpath_help('//book[title = "Oberon\'s Revenge"]/author/text()')
```

What are the titles of the books that sell for the lowest price? (Use the min aggregation function)

```{r}
xpath_help('//book[price = min(//book/price)]/title/text()')
```

List the title and prices of all books in the "Computer" genre.

* If the R `XML` package supported xpath 2.0, I would use `//book[genre = "Computer"]/concat(title, ",", price)`

```{r}
xpath_help('//book[genre = "Computer"]/*[self::title or self::price]/text()')
```

How many books cost more than $50? (Use the count aggregation function)

```{r}
xpath_help("count(//book[price > 50])")
```

## Question 2

**How many books are in each genre and what is the average price for the books in each genre? List the genre, number of books, and average price.**

```{r}
# retrieving list of all genres
genres <- xpath_help("//book/genre/text()") |> 
  sapply(function(g) as(g, "character")) |>
  unique()

get_genre <- function(genre) {
  # retrieving number of books
  numBooks <- xpath_help("count(//book[genre = '", genre, "'])")
  # retrieving average price
  avgPrice <- xpath_help("//book[genre = '", genre, "']/price/text()") |>
    sapply(function(p) as(p, "numeric")) |>
    mean()
  
  list(genre, numBooks, avgPrice)
}


do.call(rbind.data.frame, lapply(genres, get_genre)) |>
  setNames(c("Genre", "Number of Books", "Average Price"))
```




