---
title: "Indexing in SQLite"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Lesson link](http://artificium.us/lessons/60.dbdesign/l-60-611-indexing-sqlite/l-60-611.html)

Indexing can speed up data retrievals (SELECT), improve sorting (ORDER BY), and enforce uniqueness constraints (UNIQUE). Most databases automatically index the primary key -- SQLite doesn't, but MySQL does. 

Several data structures may be used for indexing: hashmaps and trees

**B-trees (SQLite)**:

* Self-balancing tree data structure
* maintains data nodes in sorted order -- searches in $O(\log _n)$ time. 
* usually > 2 children per nodes
* best for relatively large block sorting 
* during access operations entire index is in memory, but data remains on disk. 

```{r}
library(RSQLite)
db <- dbConnect(SQLite(), "MediaDB.db")
```

## Creating an index

Use `CREATE INDEX` to create an index.The general form:

```{sql eval = FALSE}
CREATE [UNIQUE] INDEX index_name [ASC|DESC] ON table_name(column_list);
```

* critical for non-prime columns when columns are frequently in joins or searches -- B-trees make equality and range searching much quicker 
* searches on a compound key should have a compound index 
* `create unique index` detects duplicated values and is generally implemented with a hashmap/hashtable
* indexes are not useful for matching searches using `LIKE`
* indexes common for `ORDER BY` -- in-order traversal of B-tree results in ordered set

```{sql connection = db}
CREATE INDEX custLN 
    ON customers(LastName);
```

## Using an Index

Indexes most useful when direct match search is done on an indexed column in the `WHERE` clause.

* relational databases automatically use an index when available/use improves query performance 

```{sql connection = db}
SELECT LastName, FirstName
  FROM customers
 WHERE LastName = 'Johansson'
```

`EXPLAIN QUERY PLAN` makes index use explicit and shows that the column is searched.

```{sql connection = db}
EXPLAIN QUERY PLAN 
SELECT LastName, FirstName
  FROM customers
 WHERE LastName = 'Johansson'
```

When the index is not useful, it is not used. 

```{sql connection = db}
EXPLAIN QUERY PLAN 
SELECT LastName, FirstName
  FROM customers
 WHERE LastName LIKE 'Johan%'
```

**Note**: SQLite does not automatically index the primary key, but does build a hashmap to quickly determine uniqueness.

```{sql connection = db}
EXPLAIN QUERY PLAN
SELECT LastName, FirstName
  FROM customers
 WHERE CustomerId = 37
```

## Deleting an Index

Use `DROP INDEX` to delete an index. General form: 

```{sql eval = FALSE}
DROP INDEX [IF EXISTS] index_name;
```

```{sql connection = db}
DROP INDEX custLN;
```

## Listing Indexes

To list all created and available indexes from a table: 

```{sql connection = db}
PRAGMA list_indexes('customers');
```

To list all indexes in a database, query the *sqlite_master* table:

```{sql connection = db}
SELECT `type`, `name`, `tbl_name`, `sql`
  FROM sqlite_master
WHERE `type` = 'index';
```

## Considerations

* searching with substring matching / `LIKE` cannot use indexes and should be avoided. 
* when new data is inserted/updated/deleted, indexes must be rearranged. The worst case run-time behaviour of updating a B-Tree is $O(n^2)$, reducing performance. 
  * indexes should be dropped prior to bulk loading as they reduce performance. After ETL they can be created again. 
* searching on expressions does not use indexes unless an [index on the expression](https://www.sqlite.org/expridx.html) is created.













