---
title: "Query a Database"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Omi Johnson

<br>

```{r}
library(RSQLite)
db <- dbConnect(SQLite(), "MediaDB.db")
```

## Queries

**(5 pts) What are the last name, city, and state of all customers who made at least one purchase and live in Brazil? Do not list duplicates.**

*Simple Approach *

```{sql eval = FALSE}
SELECT LastName, City, State
  FROM customers
 WHERE LOWER(Country) LIKE 'brazil'
   AND CustomerID IN (SELECT DISTINCT CustomerID FROM invoices);
```

*View Approach*

* I decided to create a view for all customers who have made a purchase, since this information is also desired for two other queries. 

```{sql connection = db}
CREATE VIEW IF NOT EXISTS purchasing_customers AS 
SELECT *
  FROM customers
 WHERE CustomerId IN (SELECT DISTINCT CustomerId
                        FROM invoices);
```

```{sql connection = db}
SELECT LastName, City, State
  FROM purchasing_customers
 WHERE LOWER(Country) LIKE '%brazil%'
```

<br>

**(5 pts) What are the titles of each album and the number of tracks on each album? **

```{sql connection = db}
SELECT a.Title, COUNT(DISTINCT t.TrackId) as num_tracks
  FROM ALBUMS a LEFT JOIN tracks t ON (a.AlbumId = t.AlbumId)
 GROUP BY a.AlbumID
```

<br>

**(5 pts) List the number of tracks by media type.**

```{sql connection = db}
SELECT m.Name, COUNT(DISTINCT t.TrackId) as num_tracks 
  FROM media_types m LEFT JOIN tracks t ON (t.MediaTypeId = m.MediaTypeId)
 GROUP BY m.MediaTypeId
```

<br>

**(10 pts) What is the average age of the employees? Use SQL to round the result to one digit of precision, e.g., 23.4.**

```{sql connection = db}
/* dividing by 365.25 to account for leap years */
SELECT ROUND(AVG((JULIANDAY('now') - JULIANDAY(BirthDate))/365.25), 1)
    AS average_age
  FROM employees
```

<br>

**(10 pts) What are the total number of unique customers for each US state (e.g., MA, CT, etc.), ordered alphabetically by state, who made at least one purchase (i.e., do not consider customers who never made a purchase)?**

Note: Unique customers are ensured because I am selecting from the `customers` table, where each customer is unique. 

```{sql connection = db}
SELECT State, Count(*) as distinct_customers
  FROM purchasing_customers
 WHERE Country LIKE 'USA'
 GROUP BY State
 ORDER BY State
```

<br>

**(10 pts) How many customers never made a purchase, i.e., they appear in the customers table but not in the invoices table.**

```{sql connection = db}
SELECT COUNT(*) AS num_customers_nopurchase
  FROM customers
 WHERE CustomerID NOT IN (SELECT CustomerID FROM purchasing_customers)
``` 

<br>

**(10 pts) How many albums contain the substring "symphony" in the album title?**

```{sql connection = db}
SELECT Count(*) AS num_symphony_albums
  FROM albums
 WHERE LOWER(Title) LIKE '%symphony%'
```

<br>

**(15 pts) What is the total "time" of all tracks per artist, i.e., add up the "Milliseconds" column in tracks? Display the artist's name and the time in hours rounded to two decimals. Only consider artists who published at least one track.**

* Creating a view containing merged information on artists and select track attributes, as this information is desired for the rest of the questions

```{sql connection = db}
CREATE VIEW IF NOT EXISTS artists_tracks AS
SELECT a.*, t.trackId, t.GenreId, t.Milliseconds   
  FROM (artists a LEFT JOIN albums al ON (a.ArtistId = al.ArtistId))
                  LEFT JOIN tracks t ON (al.AlbumId = t.AlbumId)
```

```{sql connection = db}
SELECT Name, ROUND(SUM(Milliseconds)/3600000.00, 2) AS summed_time_hrs
  FROM artists_tracks
 WHERE TrackId NOT NULL
 GROUP BY ArtistId
```

<br>

**(15 pts) Which artist(s) published the least number of tracks, regardless of type and format? List the artist(s) name(s). Do not list duplicates.**

```{sql connection = db}
WITH track_counts AS (
  SELECT Name, Count(DISTINCT TrackId) AS summed_count
    FROM artists_tracks
   GROUP BY ArtistId
) 
SELECT Name
  FROM track_counts
 WHERE summed_count = (SELECT MIN(summed_count) 
                         FROM track_counts)
```

<br>

**(15 pts) Which artists have composed tracks in more than three different genres? List the artists name and the number of distinct genres they have compositions in. Do not list duplicates.**

```{sql connection = db}
SELECT Name, COUNT(DISTINCT GenreId) AS genre_count
  FROM artists_tracks
 GROUP BY ArtistId
HAVING Count(DISTINCT GenreId) > 3
```

## Disconnect

```{r}
dbDisconnect(db)
```