---
title: "Transactions in SQLite"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[lesson link](http://artificium.us/lessons/70.sql/l-70-821-txns-sqlite/l-70-821.html)

A **transaction** is a sequence of database modifications and queries occuring as a single sequence. 

* the database meets all integrity constraints before and after transaction
* transactions must be explictly started and committed -- if operation within transaction fails, it must be aborted/rolled back, reversing all changes 

```{r}
library(RSQLite)
db <- dbConnect(SQLite(), "txns_demoDB.sqlitedb")
```

## SQLite and ACID

SQLite is a transactional database because it ensures that all modifications/queries exhibit **ACID properties**: atomic, consistent, isolated, and durable. 

* adherance to ACID implies that all SQLite modifications leave database in consistent state even if transaction is interrupted by program crash

### Acid Properties

**Atomic**: A sequence of modification operations cannot be broken down into smaller ones. When transaction committed, either all mods applied or none applied (no partial updates).

**Consistent**: must ensure that database is changed from one valid state to another, *e.g.* when transferring funds the total value in both accounts is the same at the beginning and end of transaction.

**Isolated**: The intermediate state of a transaction is invisible to other transactions -- concurrent transactions appear serialized. 

**Durable**: When transaction successful, changes must be permanent regardless of condition, eg. crash. On the reverse, if crash occurs mid-transaction change should not appear. 

## Transactions in SQLite

SQLite defaults to *auto-commit mode* -- for each modding command, SQLite starts transaction, processes command, and commits.

Start transactions explicitly with `BEGIN TRANSACTION`.

* SQLite does not support nested transactions
* transactions are active until explicitly committed or rolled back 
  * any changes made before committing are only visible to application initializing transaction.
  
Commit transactions with `COMMIT TRANSACTION`.

Rollback transactions with `ROLLBACK TRANSACTION`. 

* rollbacks return database to state just prior to start of transaction

## Nested Transactions

Nested transactions are useful when a change within a transaction should be made permanent while others shouldn't, all while prohibiting viewing of partial updates. 

* not supported in SQLite, but are supported in Oracle, Microsoft SQL Server

## Read Transactions

It's not common to have *SELECT* statements within transactions, but sometimes useful when guaranteeing that we read all tables from a defined state. Especially useful when generating reports.

## Example: Banking transfers

The classic transaction example is moving \$ from one account to another. 

* no client should see one account debited before other credited
* most applications log transactions into a General Journal

```{sql connection = db}
DROP TABLE IF EXISTS accounts;
```
```{sql connection = db}
DROP TABLE IF EXISTS journal;
```
```{sql connection = db}
CREATE TABLE IF NOT EXISTS accounts ( 
    account_no INTEGER NOT NULL, 
    balance DECIMAL NOT NULL DEFAULT 0,
    PRIMARY KEY(account_no),
  CHECK(balance >= 0)
);
```
```{sql connection = db}
CREATE TABLE IF NOT EXISTS journal (
    change_no INTEGER PRIMARY KEY AUTOINCREMENT,
    account_no INTEGER NOT NULL, 
    flag TEXT NOT NULL, 
    amount DECIMAL NOT NULL, 
    change_date DATE NOT NULL,
    FOREIGN KEY (account_no) 
     REFERENCES accounts(account_no)
);
```
```{sql connection = db}
INSERT INTO accounts 
     (account_no, balance)
     VALUES (117860, 1100),
            (299871, 2200);
```

Let's initiate a transfer -- moving 100 from account 1 to account 2. Requires 4 UPDATES -- updating balance in two accounts and an insertion into the journal. 

```{sql connection = db}
BEGIN TRANSACTION;
```
```{sql connection = db}
UPDATE accounts
   SET balance = balance - 100
 WHERE account_no = 117860;
```
```{sql connection = db}
UPDATE accounts
   SET balance = balance + 100
 WHERE account_no = 299871;
```
```{sql connection = db}
INSERT INTO journal
       (account_no, flag, amount, change_date)
VALUES (299871, 'C', 100, date('now'));
```
```{sql connection = db}
INSERT INTO journal
       (account_no, flag, amount, change_date)
VALUES (117860, 'D', 100, date('now'));
```
```{sql connection = db}
COMMIT;
```

Evaluating whether it worked:

```{sql connection = db}
SELECT account_no, balance FROM accounts;
```
```{sql connection = db}
SELECT * FROM journal;
```

### Transactions in Applications

In applications, we should use conditional statements t determine when/whether to commit or rollback. Within R this is best done inside a function. 

```{r}
doTransfer <- function (dbcon, acctNoSrc, acctNoDest, amount)
{
  txnFailed = FALSE
  
  if (acctNoSrc == acctNoDest)
  {
    # can't transfer to same account
    return (FALSE)
  }
  
  dbExecute(dbcon, "BEGIN TRANSACTION")

  # debit source account
  sql <- "UPDATE accounts
             SET balance = balance - ?
           WHERE account_no = ?"
  ps <- dbSendStatement(dbcon, sql,
                        params = list(amount, acctNoSrc))
  if (dbGetRowsAffected(ps) < 1)
    txnFailed = TRUE
  dbClearResult(ps)

  # credit destination account
  sql <- "UPDATE accounts
             SET balance = balance + ?
           WHERE account_no = ?"
  ps <- dbSendStatement(dbcon, sql,
                        params = list(amount, acctNoDest))
  if (dbGetRowsAffected(ps) < 1)
    txnFailed = TRUE
  dbClearResult(ps)
  
  # add accounting transaction to log
  sql <- "INSERT INTO journal 
           (account_no,flag,amount,change_date) 
          VALUES (?, 'D', ?, date('now'))"
  ps <- dbSendStatement(dbcon, sql,
                        params = list(acctNoSrc, amount))
  if (dbGetRowsAffected(ps) < 1)
    txnFailed = TRUE
  dbClearResult(ps)
  
  sql <- "INSERT INTO journal
           (account_no,flag,amount,change_date) 
          VALUES (?, 'C', ?, date('now'))"
  ps <- dbSendStatement(dbcon, sql,
                        params = list(acctNoDest, amount))
  if (dbGetRowsAffected(ps) < 1)
    txnFailed = TRUE
  dbClearResult(ps)


  # commit transaction if no failure, otherwise rollback
  if (txnFailed == TRUE)
    dbExecute(dbcon, "ROLLBACK TRANSACTION")
  else
    dbExecute(dbcon, "COMMIT TRANSACTION")

  # return status; TRUE if successful; FALSE if failed
  return (!txnFailed)
}
```

Testing

```{r}
srcAccount <- '117860'
destAccount <- '299871'
amount <- 99

status <- doTransfer(db, srcAccount, destAccount, amount)

if (status == TRUE) {
  cat('Transfer successful')
} else {
  cat('Transfer failed')
}
```
```{r}
srcAccount <- '117860'
destAccount <- '999999'
amount <- 200

status <- doTransfer(db, srcAccount, destAccount, amount)

if (status == TRUE) {
  cat('Transfer successful')
} else {
  cat('Transfer failed')
}
```
```{r}
dbDisconnect(db)
```

